Reproducible Builds for\\ Computational Research Papers

### 2020-08-22 ms.tex
improve text

\begin{abstract}
	Previous works in research reproducibility provided frameworks for writing reproducible research papers, however without considering the extra work that needs to be done by the author to make his paper reproducible and by the reviewer to verify the paper's reproducibility.
	We propose a simple template for writing computational research papers in \LaTeX\ that are easily verifiable in terms of their reproducibility.
	The template contains a Makefile that allows the verifier to execute the code that reproduces the figures, tables and variables (results) of the paper, which are then used during the \LaTeX\ compilation, to produce a pdf.
	The previous procedure is completed twice and the paper is verified as reproducible, if and only if the pdf files are identical.
	The two builds are needed to enable comparison between different builds of the same source code such that any non-deterministic stochasticity would produce different pdf files, thus making the paper non-reproducible.
	The reproducibility of a paper can be verified either locally on the machine of the verifier or remotely using Virtual Machines (VMs) in the cloud.
	We release an open source template of \textit{reproducible builds} (RBs) and use it to write and verify the reproducibility of this paper.
\end{abstract}

@@

We propose a template that uses the concept of \textit{reproducible builds} (RBs) to verify the reproducibility of computational research papers written in \LaTeX.
The template consists of a Makefile that allows the verifier to execute the source code that produces the figures, tables, variables (results), and later compile the \LaTeX\ to produce a pdf.
The previous procedure is executed twice and the research paper is verified as reproducible, if and only if the resulting pdf files are identical.
This procedure can be executed either locally in the verifier machine or remotely in Virtual Machines (VMs) in the cloud.

We provide an open source implementation of the template\footnote{\url{https://github.com/pbizopoulos/cookiecutter-reproducible-builds-for-computational-research-papers}} for writing papers with RBs and use it to write this paper\footnote{\url{https://arxiv.org/abs/2005.12660}} and verify its reproducibility\footnote{\url{https://github.com/pbizopoulos/reproducible-builds-for-computational-research-papers}}.
We use Python~\cite{van2007python} for producing the results, however other scientific-oriented programming languages such as R~\cite{ihaka1996r} and Julia~\cite{bezanson2017julia} can be used.

For the rest of the paper we will refer to:
\begin{itemize}
	\item \textit{paper} as a computational research paper,
	\item \textit{results} as the figures, tables and variables that are shown in the paper,
	\item \textit{\LaTeX\ code} as the files containing the main text (\textit{*.tex} files) and bibliography (\textit{*.bib} files) of the research paper,
	\item \textit{results code} as the files containing the scientific-oriented programming language code (\textit{*.py} files) that produces the results and
	\item \textit{code} as both the \textit{\LaTeX\ code} and \textit{results code}.
	\item \textit{author} as the author or authors of the \textit{paper},
	\item \textit{reviewer} as the reviewer of the \textit{paper},
\end{itemize}

@@

\begin{figure}[!t]
	\centering
	\begin{tikzpicture}[%
			grow via three points={one child at (0.5,-0.7) and two children at (0.5,-0.7) and (0.5,-1.4)},
			edge from parent path={(\tikzparentnode.south) |- (\tikzchildnode.west)}]
			\node[draw](root){root}
			child{node[draw, anchor=west](makefile){Makefile}}
			child{node[draw, anchor=west, fill=red!20](tex){ms.tex}}
			child{node[draw, anchor=west, fill=red!20](bib){ms.bib}}
			child{node[draw, anchor=west, fill=blue!20](py){main.py}}
			child{node[draw, anchor=west, fill=blue!20](venv){.venv}}
			child{node[draw, anchor=west, fill=blue!20](dockerfile){Dockerfile}};
			\begin{scope}[on background layer]
				\node[draw, rounded corners, dashed, fill=gray!30, inner sep=4pt, fit=(root) (dockerfile)]{};
				\node[draw, rounded corners, dashed, fill=gray!70, inner sep=2pt, fit=(py) (dockerfile)](code){};
				\node[draw, rounded corners, dashed, fill=gray!70, inner sep=2pt, fit=(tex) (bib)](paper){};
			\end{scope}
			\node[draw, right=2cm of code, fill=white] (results) {results};
			\path[draw, ->] (code) -- node[above]{venv/docker} (results);
			\node[draw, right=4cm of paper, fill=white] (pdf) {pdf};
			\path[draw, ->] (results) |- (pdf);
			\path[draw, ->] (paper) -- node[above]{docker} (pdf);
	\end{tikzpicture}
	\caption{The RBs template file hierarchy is depicted in the light gray background and the arrows denote the data flow towards generating the pdf.
	Blue indicates a \textit{results code} file, red a \textit{\LaTeX\ code} file and within the dark gray background files that are used for a subsequent step.}
	\label{fig:filehierarchyworkflow}
\end{figure}

The proposed template can be applied on computational research papers written in \LaTeX~\cite{lamport1994latex} and the \textit{results code} can be executed within a Docker container or a virtual environment.
\LaTeX\ is a typesetting system for publishing high quality research papers, and Docker~\cite{merkel2014docker} is a container system that allows portable execution of a codebase between different operating systems/environments and has shown great promise in reproducible research~\cite{boettiger2015introduction}.
The corresponding Dockerfile can be provided by the \textit{author} or generated by tools such as containerit~\cite{nust2019containerit}.

A specific file hierarchy (shown in Fig.~\ref{fig:filehierarchyworkflow}) is used in this paper however this is optional, as the template only needs to know the directories of the main files of the \LaTeX\ and \textit{results code} during its creation by the template.
The template fits in the research writing and reproducibility verificatrion in the following way:
\begin{enumerate}
	\item the \textit{author} writes or edits code in his local repository,
	\item the \textit{author} may verify the reproducibility of the paper locally,
	\item the \textit{author} may push the change to the remote repository, which optionally triggers the remote reproducibility verification:
		\begin{enumerate}
			\item two VMs (builders) execute the \textit{results code},
			\item the results along with the \textit{\LaTeX\ code} are used to compile the pdf files,
			\item the pdf files are sent to another VM (verifier), which verifies the research paper as reproducible if and only if the pdf files are identical.
		\end{enumerate}
	\item the \textit{reviewer} can check the remote reproducibility verifcation or if not available locally verify the reproducibility in his own machine.
\end{enumerate}

Docker usage in RBs provides portability of the environment of the research while the VM provides trust in the procedure (in contrast in executing these local, thus having low credibility in reporting reproducibility).
The reason that two independent builder VMs are needed is to enable comparison between different builds of the same source code and any non-deterministic stochasticity (such as not applying a specific seed to random generators) would produce different pdfs, thus making the verifier VM report the paper as non-reproducible.

Additional constraints could be imposed in the VMs such as disallowing image files in the code and monitoring the network connections of the \textit{code} to help the \textit{reviewer} check whether the results are generated from the \textit{code} and are not fetched from an external source.

Other potential uses of the RBs template include:
\begin{itemize}
	\item Preprint server (such as arXiv) results reproducibility verification.
		The difference with the remote reproducibility verification is the presence of an additional VM that checkouts the \textit{\LaTeX\ code} and generates a pdf which is then compared with the one generated from \textit{code}.
	\item Regression testing and debugging, to ensure that changes to the code do not alter the results.
\end{itemize}

@@

Regarding research that is computational costly, the template provides a debug flag that verifies the reproducibility of a computationally lighter version of the paper.
This flag also helps in faster iterations during research writing and experimentation.
An example of the debug flag in the field of neural networks is setting the number of epochs and number of training samples to a low value.

Some possible executions that RBs provide are summarized in the following syntax:
\begin{lstlisting}[language=Bash, style=lststyle, caption={Makefile call syntax from the shell.}, captionpos=b]

@@

\section{Discussion}
Previous works in this area such as Hurlin et al.~\cite{hurlin2019reproducibility} propose an `external certification agency' for economic studies.
RBs provide a simple solution for writing and verifying the reproducibility of a computational research paper as a whole, without having to trust a third-party.
Anyone can clone a repository written with RBs and locally verify its reproducibility, in case third party verifications (such as cloud VMs) cannot be trusted.

\section{APPENDIX - Improving Computational Research Reproducibility}
This appendix provides suggestions for improving the reproducibility of computational research papers written in \LaTeX\ with Python as the \textit{results code} programming language.
For demonstration purposes we train and test a simple neural network on the MNIST~\cite{lecun2010mnist} dataset with PyTorch~\cite{paszke2019pytorch}.

A useful \LaTeX\ package for automatically embedding code results variables in \textit{\LaTeX\ code} is \textit{datatool} with the following example use:
\begin{lstlisting}[language=TeX, style=lststyle, caption={\LaTeX\ datatool example of loading a file that contains pairs of keys and values (keys\_values.csv) generated by a \textit{results code} and getting the value of a key named lr.}, captionpos=b]
\DTLloaddb{keys_values}{keys_values.csv}
\DTLfetch{keys_values}{key}{lr}{value}
\end{lstlisting}

@@

Another useful results function for Python is \textit{pandas.DataFrame.to\_latex} which automatically converts a dataframe table to a \LaTeX\ table (as shown in Table~\ref{table:table}.

@@

\begin{table}[h]
	\centering
	\caption{Table example created from \textit{results code}.}
	\label{table:table}
	\setlength\tabcolsep{4.2pt}
	\input{results/table.tex}
\end{table}

>>>>

\begin{abstract}
	Previous works in research reproducibility provided frameworks for writing reproducible research papers, however without considering the cost of extra work that needs to be done by the author to make his paper reproducible and by the reviewer to verify the paper's reproducibility.
	We propose a simple template for writing computational research papers in \LaTeX\ that are easily verifiable in terms of their reproducibility.
	The template contains a Makefile that allows the reviewer to execute the code that reproduces the figures, tables and variables (results) of the paper, which are then used during the \LaTeX\ compilation, to produce a pdf.
	The previous procedure is completed twice and the paper is verified as reproducible, if and only if the pdf files are identical.
	The two builds are needed to enable comparison between different builds of the same source code such that any non-deterministic stochasticity would produce different pdf files, thus making the paper non-reproducible.
	The reproducibility of a paper can be verified either locally on the machine of the reviewer or remotely using Virtual Machines (VMs) in the cloud.
	We release an open source template of \textit{reproducible builds} (RBs) and use it to write and verify the reproducibility of this paper.
\end{abstract}

@@

We propose a template that uses the concept of \textit{reproducible builds} (RBs) to verify the reproducibility of computational research papers written in \LaTeX in a local or remote computational environment.
We provide an open source implementation of the template\footnote{\url{https://github.com/pbizopoulos/cookiecutter-reproducible-builds-for-computational-research-papers}} for writing papers with RBs and use it to write this paper\footnote{\url{https://arxiv.org/abs/2005.12660}} and verify its reproducibility\footnote{\url{https://github.com/pbizopoulos/reproducible-builds-for-computational-research-papers}}.
We use Python~\cite{van2007python} for producing the results, however other scientific-oriented programming languages can be used.

For the rest of the paper we will refer to:
\begin{itemize}
	\item \textit{paper} as a computational research paper,
	\item \textit{results} as the figures, tables and variables that are shown in the \textit{paper},
	\item \textit{\LaTeX\ code} as the files containing the main text (\textit{*.tex} files) and bibliography (\textit{*.bib} files) of the \textit{paper},
	\item \textit{results code} as the files containing the scientific-oriented programming language code (\textit{*.py} files) that produces the \textit{results},
	\item \textit{code} as both the \textit{\LaTeX\ code} and \textit{results code},
	\item \textit{author} as the author(s) of the \textit{paper} and
	\item \textit{reviewer} as the reviewer(s) of the \textit{paper} or anyone in interest of verifying the \textit{paper} (author(s), journal editors, other researchers).
\end{itemize}

@@

\begin{figure}[!t]
	\centering
	\begin{tikzpicture}[%
			grow via three points={one child at (0.5,-0.7) and two children at (0.5,-0.7) and (0.5,-1.4)},
			edge from parent path={(\tikzparentnode.south) |- (\tikzchildnode.west)}]
			\node[draw](root){root}
			child{node[draw, anchor=west](makefile){Makefile}}
			child{node[draw, anchor=west, fill=red!20](tex){ms.tex}}
			child{node[draw, anchor=west, fill=red!20](bib){ms.bib}}
			child{node[draw, anchor=west, fill=blue!20](py){main.py}}
			child{node[draw, anchor=west, fill=blue!20](venv){.venv}}
			child{node[draw, anchor=west, fill=blue!20](dockerfile){Dockerfile}};
			\begin{scope}[on background layer]
				\node[draw, rounded corners, dashed, fill=gray!30, inner sep=4pt, fit=(root) (dockerfile)]{};
				\node[draw, rounded corners, dashed, fill=gray!70, inner sep=2pt, fit=(py) (dockerfile)](code){};
				\node[draw, rounded corners, dashed, fill=gray!70, inner sep=2pt, fit=(tex) (bib)](paper){};
			\end{scope}
			\node[draw, right=2cm of code, fill=white] (results) {results};
			\path[draw, ->] (code) -- node[above]{venv/docker} (results);
			\node[draw, right=4cm of paper, fill=white] (pdf) {pdf};
			\path[draw, ->] (results) |- (pdf);
			\path[draw, ->] (paper) -- node[above]{docker} (pdf);
	\end{tikzpicture}
	\caption{The RBs template file hierarchy is depicted in the light gray background and the arrows denote the data flow towards generating the pdf.
	Blue indicates a \textit{results code} file, red a \textit{\LaTeX\ code} file and within the dark gray background, files that are used for a subsequent step.}
	\label{fig:filehierarchyworkflow}
\end{figure}

The proposed template can be applied on \textit{papers} written in \LaTeX~\cite{lamport1994latex} and the \textit{results code} can be executed within a Docker container or a virtual environment.
\LaTeX\ is a typesetting system for publishing high quality research papers, and Docker~\cite{merkel2014docker} is a container system that allows portable execution of a codebase between different operating systems/environments and has shown great promise in reproducible research~\cite{boettiger2015introduction}.
The corresponding Dockerfile can be provided by the \textit{author} or generated by tools such as containerit~\cite{nust2019containerit}.

A specific file hierarchy (shown in Fig.~\ref{fig:filehierarchyworkflow}) is used in this paper however this is optional, as the template only needs to know the directories of the main files of the \LaTeX\ and \textit{results code} during its creation by the template.
The template fits in the research writing and reproducibility verification in the following way:
\begin{enumerate}
	\item the \textit{author} edits the \textit{code} in the local repository,
	\item the \textit{author} may verify the reproducibility of the \textit{paper} locally:
		\begin{enumerate}
			\item the \textit{results code} is executed twice in the local machine,
			\item the \textit{results} along with the \textit{\LaTeX\ code} are used to compile the pdf files,
			\item the \textit{paper} is verifed as locally reproducible if and only if the pdf files are identical.
		\end{enumerate}
	\item the \textit{author} may push the change to the remote repository, which optionally triggers the remote reproducibility verification:
		\begin{enumerate}
			\item the \textit{results code} is executed twice in two cloud VMs (builders),
			\item the \textit{results} along with the \textit{\LaTeX\ code} are used to compile the pdf files,
			\item the \textit{paper} is verified as remotely reproducible if and only if the pdf files are identical.
		\end{enumerate}
	\item the \textit{reviewer} may verify the reproducibility of the paper on a local or remote machine.
\end{enumerate}

Docker usage in RBs provides portability of the environment of the research while the VM provides trust in the procedure (in contrast with verifying locally, thus having low credibility in reporting reproducibility).
The reason that two local/remote compilations are needed is to enable comparison between different builds of the same source code and any non-deterministic stochasticity (such as not applying a specific seed to random generators) would produce different pdfs, thus making the \textit{paper} non-reproducible.

Additional constraints could be imposed in the VMs such as disallowing image files in the \textit{code} and monitoring the network connections of the \textit{code} to help the \textit{reviewer} check whether the \textit{results} are generated from the \textit{code} and are not fetched from an external source.

Other potential uses of the RBs template include:
\begin{itemize}
	\item Preprint server (such as arXiv) \textit{paper} reproducibility verification.
		The difference with the remote reproducibility verification is the presence of an additional VM that checkouts the \textit{\LaTeX\ code} and generates a pdf which is then compared with the one generated from \textit{code}.
	\item Regression testing and debugging, to ensure that changes to \textit{code} do not alter the results.
\end{itemize}

@@

Regarding research that is computational costly, the template provides a debug flag that verifies the reproducibility of a computationally lighter version of the \textit{paper}.
This flag also helps in faster iterations during research writing and experimentation.
An example of the debug flag in the field of neural networks is setting the number of epochs and number of training samples to a low value.

Executing RBs can be done using the following:
\begin{lstlisting}[language=Bash, style=lststyle, caption={Makefile call syntax from the shell.}, captionpos=b]

@@

\section{Example Use Case of RBs}
This sections provides a use case of RBs and also serves as a manual for creating reproducibility of \textit{papers} written in \LaTeX\ with Python as the \textit{results code} programming language.
For demonstration purposes we train and test a simple neural network on the MNIST~\cite{lecun2010mnist} dataset with PyTorch~\cite{paszke2019pytorch}.

A useful \LaTeX\ package for automatically embedding variables from \textit{results} in \textit{\LaTeX\ code} is \textit{datatool} with the following example use:
\begin{lstlisting}[language=TeX, style=lststyle, caption={\LaTeX\ datatool example of loading a file that contains pairs of keys and values (keys\_values.csv) generated by a \textit{results code} and getting the value of a key named lr.}, captionpos=b]
\DTLloaddb{keys_values}{keys_values.csv}
\DTLfetch{keys_values}{key}{lr}{value}
\end{lstlisting}

@@

Another useful function for Python is \textit{pandas.DataFrame.to\_latex} which automatically converts a dataframe table to a \LaTeX\ table (as shown in Table~\ref{table:table}.

@@

\begin{table}[h]
	\centering
	\caption{Table example created from results code.}
	\label{table:table}
	\setlength\tabcolsep{4.2pt}
	\input{results/table.tex}
\end{table}

\section{Discussion}
Previous works in this area such as Hurlin et al.~\cite{hurlin2019reproducibility} propose an `external certification agency' for economic studies.
RBs provide a simple solution for writing and verifying the reproducibility of a \textit{paper} as a whole, without having to trust a third-party.
Anyone can clone a repository written with RBs and locally verify its reproducibility, in case third party verifications (such as cloud VMs) cannot be trusted.
Similar to RBs is ReDoc~\cite{schwab2000making} which however focuses on separate result file reproducibility and requires the author to write each rule in the Makefile for each result file.
RBs on the other hand focus on `~\textit{paper} as a whole' reproducibility with an automatically generated Makefile.
